---
title: "Introduction to R"
author: "Clay Ford"
date: "Summer 2022"
output: slidy_presentation
---

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(echo = TRUE, prompt = TRUE, comment="",
                      tidy.opts = list(width.cutoff = 50), tidy = TRUE)
```

## What's the plan?

- Learn essential R concepts.
- Perform some basic data analysis and graphics.
- Lay a foundation to go further.

R is big and powerful and takes time to learn. But commit and do the work and you will be rewarded!

## R plus RStudio

- RStudio is an interactive development editor (IDE) _that runs R for you_.
- You need to install both, but you only need to open RStudio.
- Opening RStudio also opens R.
- RStudio makes using and learning R easier (in my opinion)

## The R Console

- Opening R/RStudio begins an _R session_.
- The Console is the R program.
- We enter _expressions_ at the Console prompt (`>`), hit Enter, and R _evaluates_ the expression and returns output.
- For example, the Console can be used as a calculator:

```{r}
22/7
```

R evaluates the mathematical expression and returns the answer. Math operators include `+`, `-`, `*`, `/`, `^` (exponent), `%%` (division remainder), and `%/%` (division quotient)

## Assignment

- If we want to _save_ the output, we assign it a name.
- To do this we use the assignment operator: `<-` 
- Keyboard shortcut: `Alt + -` (Win/Linux) or `Option + -` (Mac).
- A name must begin with a letter and not contain spaces.
- For example, save our calculation as "Pi":

```{r}
Pi <- 22/7
```

R evaluates the expression and assigns it the name `Pi`. `Pi` is now an _object_ in memory. R is case sensitive. "Pi" and "pi" are different. (`pi` is a built-in constant: 3.141593)
 
Note: You can also use `=` but most style guides recommend `<-`. There are subtle differences that go beyond the scope of this workshop.

## Functions

- R uses _functions_ to work with objects.
- Functions take _arguments_ inside parentheses.
- Example: the `sqrt` function takes the square root of a numeric object. In this example, the `Pi` object is the argument.


```{r}
sqrt(Pi)
```

Another function is `round`. Notice the second argument allows us to specify number of digits. Arguments are separated by commas.

```{r}
round(Pi, digits = 2)
```

## Data

- The main reason to use R is for the statistical analysis and visualization of data. To do that we need to _import_ data into R.
- R can import virtually any type of data file.
- A common data file is CSV. The `read.csv` function imports CSV files.
- Example: import a CSV file named `albemarle_homes.csv` and save as an object named `homes`:

```{r echo=FALSE}
homes <- read.csv("https://github.com/uvastatlab/DJA/raw/main/data/albemarle_homes_2022.csv")
```


```{r eval=FALSE}
homes <- read.csv('albemarle_homes.csv')
```


We place the file name in quotes since it's the name of a file and not a function or object.

The result is an object called `homes` that is stored in memory and available for use in R.

## Working directories

- The function `read.csv("albemarle_homes.csv")` tells R to import a CSV file named "albemarle_homes.csv". Where does it look for it?
- Answer: the _working directory_
- The working directory is where your computer looks for or saves files (assuming you don't tell it where to look).
- The current working directory is listed at the top of the Console window in RStudio.
- If the file is not in the working directory R returns an error message.

```{r error = TRUE}
homes <- read.csv('survey.csv')
```


## Setting the working directory

- You can manually set the working directory in RStudio by going to **Session...Set Working Directory...Choose Directory...**
- You can also set it using the `setwd` function. For example:
```{r eval=FALSE}
setwd("/Users/clayford/projects/evil_plan/data")
```

- `"/Users/clayford/projects/evil_plan/data"` is called a _file path_.
- Starting a path with `/` means start from your _root_ directory.
- Starting a path with `~/` means start from your _home_ directory.
- Hitting `Tab` after `/` or `~/` in RStudio will open a list of available folders and allow you to drill down to the desired directory.
- Can also use paths in functions like `read.csv`.

```{r eval=FALSE}
d <- read.csv("/Users/clayford/data/scores.csv")
```


## Data frames

- Rectangular data structures such as CSV files are read in as _data frames_.
- The columns are usually _variables_, such as gender, income, volume, mass, score, treatment, date, state, group, etc.
- The rows are usually _observations_.
- We can extract a column using the `$` operator:

```{r}
median(homes$TotalValue)
```

The `median` function calculates the median of the `TotalValue` column of the `homes` data frame.

## Vectors

- Columns of data frames are known as _vectors_. 
- Vectors contain the same _type_ of data (numeric, character, etc)
- We can create our own vectors using the `c` function.
- Many R functions are _vectorized_, meaning they work on vectors.
- Example: The `nchar` functions counts number of characters in a phrase

```{r}
nchar(c("Virginia", "Maryland", "West Virginia"))
```

We didn't have to "loop" through the values of the vector. Since `nchar` is vectorized, it did it for us.

## R scripts

- Instead of entering functions in the Console one at a time, we often will write multiple lines of _R code_ in a script.
- A script is a text file with a `.R` extension.
- When we work with scripts we can submit one or more _lines of code_ to the Console. 
- We can also save our code and share with colleagues.
- To start a new R Script, go to **File...New File...R Script**.


## Working in R scripts

- Writing code in an R script is like writing text. Hitting Enter starts a new line.
- To submit a line of R code from a script: place the cursor in the line and hit `Ctrl + Enter` (Win/Linux) or `Cmd + Return` (Mac).
- To submit a _many lines_ of R code from a script: highlight the lines and hit `Ctrl + Enter` (Win/Linux) or `Cmd + Return` (Mac).
- Any output is printed in the R console.
- RStudio allows you to have multiple R scripts open.
- Text prefaced with `#` are treated as _comments_. These are notes about the R code that are not evaluated.
- In this workshop we'll work through and review a R script I have already written: `Intro_R_2021.R`


## R packages

- Just as you can install apps on your phone, you can install _packages_ in R.
- Packages are collections of functions that serve a particular purpose.
- Example: The {readxl} package provides functions for importing Excel files.
- Packages are free to install and require an internet connection.
- In RStudio, go to **Tools...Install R packages...**, enter the package name(s) and click Install.
- To access functions in a package, use the `library` function to _load_ the package. Example: load {readxl} so we can use the `read_excel` function

```{r eval=FALSE}
library(readxl)
d <- read_excel('data.xlsx')
```

- Packages only need to be installed once, but loaded every new R session.
- To check for package updates, go to **Tools...Check for Package Updates...**

## How do I know which R packages I need?

- You will learn from colleagues, classes, books, conferences, blog posts, Twitter, Reddit, Googling for help, and so on.
- There are thousands of R packages; you will never use or need them all.
- When you install R, you get about 30 "recommended" packages collectively known as _base R_.
- To see what packages you have installed, click the Packages tab in RStudio.
- Most R packages are developed by individuals; some are well-maintained and well-documented, others not so much.

## Getting Help

- Your best bet is Googling an error message or searching for what you need help with; it's probably been answered before
- Typing `?` before a function in the console and hitting Enter will take you to a help page for that function. (eg, `?substr`)
- At the bottom there will usually be examples of how to use the function.
- The `example` function will run the examples for you so you can examine what the function does. (eg, `example(substr)`)
- You can also highlight example code in RStudio and hit `Ctrl + Enter` (Win/Linux) or `Cmd + Return` (Mac) to run the code.
- Quality of examples vary; some are terse and minimal, some are treasure troves of useful R code.

## A basic analysis workflow in R

1. Import data into R 
2. Clean or "wrangle" the data if needed
3. Analyze and visualize data

We'll walk through an example analysis in R and talk about these steps.

Steps 1 and 2 require various levels of R expertise.

Step 3 requires not just R expertise, but also statistical and subject matter knowledge.

Let's open the R script `Intro_R_2021.R`

## Keep learning R

- I encourage you to take more Introductory R classes and workshops. There are many opinions about how to teach and use R.
- Other Intro to R workshops will almost certainly emphasize different functions and "best practices".
- Subscribe to R Weekly, a round up of tutorials, blog posts and news about R: https://feedburner.google.com/fb/a/mailverify?uri=rweekly&loc=en_US
- Search YouTube for "getting started with R"


## Online tutorials and books

**Other Intro to R tutorials**

- fasteR: Fast Lane to Learning R!   
https://github.com/matloff/fasteR

- Getting Started in R: Tinyverse Edition  
https://eddelbuettel.github.io/gsir-te/Getting-Started-in-R.pdf

- The Original Intro to R   
https://cran.r-project.org/doc/manuals/r-release/R-intro.html

**Online Books**

- Modern Data Science with R  
https://mdsr-book.github.io/mdsr2e/

- R for Data Science  
https://r4ds.had.co.nz/

## Thank you

- For help and advice with R and statistical analyses: statlab@virginia.edu

- Sign up for more workshops or see past workshops:
http://data.library.virginia.edu/training/

- Register for the Research Data Services newsletter to be notified of new workshops: http://data.library.virginia.edu/newsletters/